---
title: '<img src="https://drive.google.com/uc?id=118y5T73-iSUZoAMtoJGddxq9QzD_GDKX" style="height:40px;float:left;margin:10px" /><img src="https://drive.google.com/uc?id=1fHQUzF3ZjaoHj9KQ33-94dK_X1hcmjzW" style="height:50px;float:left;margin:10px" />'
pagetitle: 'AA Trigger Report'
output:
  html_document:
    css: ../country_template/docs/style.css
    includes:
      in_header: ../country_template/docs/header.html
    toc: no
    toc_float: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    css: ../country_template/docs/style.css
    includes:
      in_header: ../country_template/docs/header.html
    toc: no
    toc_float: yes
    toc_depth: '4'
    df_print: paged
editor_options:
  chunk_output_type: console
---
<style type="text/css">
.main-container {
  max-width: 90%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r README, echo = FALSE, message = FALSE, warning = FALSE}
### NOTE: The template is set up for two 3-timepoint predictive triggers and a 1-timepoint observational trigger
### NOTE: Performance metrics and their CIs should have been generated separately before the trigger report can be produced.

### Create a trigger_report folder under DATA_DIR > /private/exploration/{country_pcode3/trigger_performance}
### Copy the dummy_learning_table_for_template.csv file to the trigger_report folder on DATA_DIR; remove 'dummy" and "for_template" from filename; add the 3-letter pcode for the country at the very beginning of the filename (eg: ner_perf_metrics_table.csv)
### Create a "plots" folder sister to the RMD file in the country's `analyses` folder
### Update the "hard-coded-values" code chunk below
### If there are fewer or more than 3 triggers and/or fewer or more than 2 packages: Adjust the number of triggers and/or Packages throughout the Rmd (review the snapshot table and activation table carefully.) 
### Review, adjust or remove the country-specific code chunk
### Remove the Annex if no trigger has more than 1 activation timepoint or adjust the Annex as needed 
### Knit
```

```{r hard-coded-values, echo = FALSE, message = FALSE, warning = FALSE}
country_name <- "Niger"
country_pcode3 <- "ner"
shock <- "Drought"
nbr_hist_shocks <- 9 # X is a number. #NER: Number of reported bad years 

# trigger A info  
name_a <- "Trigger1" # what name is the trigger listed as in the perf_metrics_table.csv files?
type_a <- 'Predictive' # Predictive or Observational?
monitored_area_a <- "National"
timepoints_a<- 'Jan, Feb, Mar' # eg: 'Jan, Feb, Mar'
package_a <- 'Package 1' # eg: 'Package 1'
funding_a <- 5.25 # input number only, no $ sign or 'm' for million. Formatting will be done automatically.
targeted_window_a <- 'Jul-Sep' # 3-letter names for first and last month. eg: 'Jul-Sep'. Forecasted period or period over which obs data are computed.
lead_time_a <- '6-4 months' # 'eg: 7-5 months', Change units as needed. Include all lead times if multiple activation timepoints
data_source_a <- 'IRI flexible seasonal forecast' # name of product + provider
hist_period_a <- "1991 - 2020" # period for which historical data is available
monitored_by_a <- 'IRI' # owner of monitoring responsibilities

# trigger B info  
name_b <- "Trigger2"
type_b <- 'Predictive'
monitored_area_b <- "National"
timepoints_b <- 'Apr, May, Jun'
package_b <- 'Package 2'
funding_b <- 9.5
targeted_window_b <- 'Jul-Sep'
lead_time_b <- '3-1 months'
data_source_b <- 'IRI flexible seasonal forecast'
hist_period_b <- "1991 - 2020"
monitored_by_b <- 'IRI'

# trigger C info  
name_c <- "Trigger3"
type_c <- 'Observational'
monitored_area_c <- "National"
timepoints_c <- 'Aug'
package_c <- 'Package 2'
funding_c <- 9.5
targeted_window_c <- 'Jun-Jul'
lead_time_c <- 'N/A'
data_source_c <- 'ENACTS or CHIRP'
hist_period_c <- "1991 - 2020"
monitored_by_c <- 'IRI'

# activation timepoints for triggers with multiple timepoints ONLY
atv_timepoints <- c("Jan", "Feb", "Mar", "Apr", "May", "June")
  
# Compile list of triggers to be included in the report
trigger_list <- c(name_a, name_b, name_c, atv_timepoints)
scenario_list <- c("framework-full", "framework-p1", "framework-p2") # exclude framework-none and framework-any

# Table footnotes
trigger_table_footnote <- "Expected cost computed using most likely probability of activation (central value). Triggers 1 and 2 can reach their threshold and activate independently from one another. Trigger 3 can only be met if Trigger 2 was not met. See Annex for performance metrics per timepoint. Trigger performance assessed by comparing recommended activations with historical bad years (i.e., years with a shock). 'Bad years' were identified through convergence of evidence from data on cereal deficit, millet production, food insecurity, fodder needs, and farmer surveys."

full_atv_definition <- "A full activation occurs when Triggers 1 and 2 are met, or when Triggers 1 and 3 are met."
package1_atv_definition <- "Package 1 is activated when Trigger 1 is met."
package2_atv_definition <- "Package 2 is activated when Trigger 2 or Trigger 3 is met."
activation_table_footnote <-  "" # additional notes, if needed

# Data providers
data_providers <- "The Département de la statistique du Niger, the Direction de la météorologie nationale du Niger, IRI Columbia University, the World Food Programme, and the Food and Agriculture Orgnization" # will be followed by "provide data and/or analytical support."

# Analysis
analysis <- "Historical data and analysis are available through IRI's Maproom [tool](http://iridl.ldeo.columbia.edu/fbfmaproom2/niger). The triggers were developed by IRI in close collaboration with DMN, WFP Niger, and the Centre for Humanitarian Data with valuable input from participating agencies and OCHA."

# Monitoring process
monitoring_process <- "Between January and June inclusively, the FIT team at IRI updates the forecast (Triggers 1 and 2) and the decision tool within Maproom by the 22nd of each month. Within 24 hours of the update, the IRI team notifies the AA team, the Niger Humanitarian Coordinator, and the Chief of CERF by email whether or not the trigger is met. By 7 August the Direction de la météorologie nationale du Niger will collate the rainfall measurements from the stations specified in the framework for the period 1 June - 31 July, and share them with the IRI FIT team. By 10 August the IRI FIT team will confirm the number of stations for which complete data is available; select ENACTS (80% or more available) or CHIRP (less than 80% available) accordingly; compute the SPI values and update Maproom. The IRI FIT team will then notify the AA team, the Niger Humanitarian Coordinator, and the Chief of CERF by email whether or not the trigger is met. In case of an activation, additional meteorological data will be shared by the DMN, the IRI FIT team, and/or Centre for Humanitarian Data to inform response targeting."
  
```

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE) # do not print code by default
knitr::opts_chunk$set(include = TRUE) # include chunk output by default
knitr::opts_chunk$set(message = FALSE) # do not print messages by default
knitr::opts_chunk$set(warning = FALSE) # do not print warnings by default

options(scipen = 999) # turn off scientific notation
options(encoding = "UTF-8") # set encoding to UTF-8 instead of ANSI
```

```{r libraries, include = FALSE, warning = FALSE}
packages <- c('tidyverse', 'ggthemes', 'kableExtra', 'knitr', 'flextable')

# install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# load libraries 
lapply(packages, library, character.only = TRUE)

```


```{r paths}
data_folder_path <- paste0(Sys.getenv("AA_DATA_DIR"), "/private/exploration/", country_pcode3,"/trigger_performance/")
perf_metrics_95_filename <- paste0(data_folder_path, country_pcode3, "_perf_metrics_table_ci_0.95.csv")
perf_metrics_68_filename <- paste0(data_folder_path, country_pcode3, "_perf_metrics_table_ci_0.68.csv")
learning_table_filename <- paste0(data_folder_path, country_pcode3, "_learning_table.csv")
```

```{r functions}
source("../country_template/docs/plotAtvCI.R")
source("../country_template/docs/plotTradeoffCI.R")
```

```{r data-loading-and-formatting}

# read in performance metrics files
perf_metrics_95_data <- read.csv(perf_metrics_95_filename,
                              colClasses = c('character', 'character', 'numeric', 'character')
                              )

perf_metrics_68_data <- read.csv(perf_metrics_68_filename,
                              colClasses = c('character', 'character', 'numeric', 'character')
                              )

# format performance metrics values
perf_metrics_95_data$value <- round(100 * perf_metrics_95_data$value, 0)
perf_metrics_68_data$value <- round(100 * perf_metrics_68_data$value, 0)

# uniquely identify point labels
perf_metrics_95_data$upoint <- paste0(perf_metrics_95_data$point, "_95")
perf_metrics_68_data$upoint <- paste0(perf_metrics_68_data$point, "_68")

# combine datasets
perf_metrics_data <- rbind(perf_metrics_95_data, perf_metrics_68_data) %>%
                      select(-point) # remove non-unique point identifier

# create & append framework-none metrics
framework_min_metrics <- perf_metrics_data %>%
                            filter(trigger == 'framework-min' & metric == 'atv') %>%
                            mutate(value = 100 - value)

framework_none_metrics <- data.frame(metric = 'atv', trigger = 'framework-none', value = NA, upoint = c('central_95', 'low_end_95', 'high_end_95', 'central_68', 'low_end_68', 'high_end_68'))

framework_none_metrics$value[framework_none_metrics$upoint == 'central_95'] <- framework_min_metrics$value[framework_min_metrics$upoint == 'central_95']
framework_none_metrics$value[framework_none_metrics$upoint == 'low_end_95'] <- framework_min_metrics$value[framework_min_metrics$upoint == 'high_end_95']
framework_none_metrics$value[framework_none_metrics$upoint == 'high_end_95'] <- framework_min_metrics$value[framework_min_metrics$upoint == 'low_end_95']
framework_none_metrics$value[framework_none_metrics$upoint == 'central_68'] <- framework_min_metrics$value[framework_min_metrics$upoint == 'central_68']
framework_none_metrics$value[framework_none_metrics$upoint == 'low_end_68'] <- framework_min_metrics$value[framework_min_metrics$upoint == 'high_end_68']
framework_none_metrics$value[framework_none_metrics$upoint == 'high_end_68'] <- framework_min_metrics$value[framework_min_metrics$upoint == 'low_end_68']

perf_metrics_data <- rbind(perf_metrics_data, framework_none_metrics)

# format funding amounts
funding_a_formatted <- paste0('$',funding_a,'m')
funding_b_formatted <- paste0('$',funding_b,'m')
funding_c_formatted <- paste0('$',funding_c,'m')

# expected cost amounts (probability of activation multiplied by funding amount)
expected_cost_a <- paste0('$', round(funding_a * (perf_metrics_data %>% filter(trigger == name_a & metric == "atv" & upoint == 'central_95') %>% select(value) %>% as.numeric()/100), 2), 'm')
expected_cost_b <- paste0('$', round(funding_b * (perf_metrics_data %>% filter(trigger == name_b & metric == "atv" & upoint == 'central_95') %>% select(value) %>% as.numeric()/100), 2), 'm')
expected_cost_c <- paste0('$', round(funding_c * (perf_metrics_data %>% filter(trigger == name_c & metric == "atv" & upoint == 'central_95') %>% select(value) %>% as.numeric()/100), 2), 'm')

```

```{r NER-specific}

hist_act <- read.csv(paste0(data_folder_path, "historical_activations_trigger_v1.csv"))
p2_pred <- sum(hist_act$Trigger2 == 'yes') # count activations of T2 (prediction-based activation of Package 2)
p2_obs <- sum(hist_act$Trigger3 == 'yes' & hist_act$Trigger2 == '') # count activations of T3 when T2 was not met (observation-based activation of Package 2)
p2_tot <- sum(hist_act$Trigger3 == 'yes' | hist_act$Trigger2 == 'yes') # count activations of P2

package2_atv_definition <- paste0(package2_atv_definition, " Historically Package 2 disbursement was triggered by the predictive trigger (Trigger 2) ", round(100*p2_pred/p2_tot, 0),"% of the time and by the observational trigger (Trigger 3) ", round(100*p2_obs/p2_tot, 0), "% of the time.")
```


```{r atv-CIs}

# Create a CI visual for atv for each trigger, activation timepoint, and activation scenario beside None and Any

for(trigger in c(trigger_list, scenario_list)) {

  # produce segment_dimensions and plots for triggers 
  output <- plotAtvCI(trigger_id = trigger)
  
  # save plot as png
  filename <- paste0(trigger, "_atv.png")
  png(filename = paste0("plots/", filename), width = 500, height = 400, units = "px")
  print(output$p)
  invisible(dev.off())
}

```

```{r tradeoff-CIs}

# Tradeoff CIs for individual triggers

for(trigger in c(trigger_list)) {
  for (metric in c('var', 'det')) {
  
  output <- plotTradeoffCI(trigger_id = trigger, 
                   left_metric_name = metric)
  
  # save plot as png
  filename <- paste0(trigger, "_", metric, "_tradeoff_bar.png")
  png(filename = paste0("plots/", filename), width = 500, height = 400, units = "px")
  print(output$p)
  invisible(dev.off())
  }
}
  
# Tradeoff CI for activation scenarios

  output <- plotTradeoffCI(trigger_id = "framework-min", 
                   left_metric_name = "atv")
  
  # save plot as png
  filename <- paste0("framework-min_atv_tradeoff_bar.png")
  png(filename = paste0("plots/", filename), width = 500, height = 400, units = "px")
  print(output$p)
  invisible(dev.off())

```

```{r file-formatting-and-saving}

# list plots to be trimmed
plots_to_trim <- grep(list.files(path = "./plots"), 
                      pattern = 'trimmed_', 
                      invert = TRUE, 
                      value = TRUE) # excludes 'trimmed_' graphs, which will be overwritten

# trim plots
for (filename in plots_to_trim) {

  # read in the image
  original_plot <- magick::image_read(paste0("plots/", filename))
  
  # trim the image
  trimmed_plot <- magick::image_trim(original_plot)
  
  # save trimmed plot
  magick::image_write(trimmed_plot, path = paste0("plots/trimmed_", filename), format = "png")

}

```

```{r key-messages-metrics}
any_atv_low_end_68 <- perf_metrics_data %>% filter(trigger == 'framework-min' & metric == 'atv') %>% filter(upoint == 'low_end_68') %>% select(value) %>% as.numeric()

any_atv_high_end_68 <- perf_metrics_data %>% filter(trigger == 'framework-min' & metric == 'atv') %>% filter(upoint == 'high_end_68') %>% select(value) %>% as.numeric()

#any_det_low_end_68 <- perf_metrics_data %>% filter(trigger == 'framework-min' & metric == 'det') %>% filter(upoint == 'low_end_68') %>% select(value) %>% as.numeric()
any_det_low_end_68 <- 89 # manual overwrite using the rule of three

#any_det_high_end_68 <- perf_metrics_data %>% filter(trigger == 'framework-min' & metric == 'det') %>% filter(upoint == 'high_end_68') %>% select(value) %>% as.numeric()
any_det_high_end_68 <- 100 # manual overwrite using the rule of three

any_far_low_end_68 <- perf_metrics_data %>% filter(trigger == 'framework-min' & metric == 'far') %>% filter(upoint == 'low_end_68') %>% select(value) %>% as.numeric()

any_far_high_end_68 <- perf_metrics_data %>% filter(trigger == 'framework-min' & metric == 'far') %>% filter(upoint == 'high_end_68') %>% select(value) %>% as.numeric()

mitigation_warranted <- ifelse(any_far_low_end_68 >= 40, "Mitigation is recommended to minimise the negative impact of a false alarm due to its moderate to high likelihood.", "")

```

# Anticipatory Action Trigger Mechanism Report
<p style="color:#007ce1;font-size:22px;text-align:center;"> `r paste0(shock, ' in ', country_name)`</p> 
<p style="color:black;font-size:14px;text-align:right;"> Last updated: `r Sys.Date()`</p> 

## Key Messages

* <p style="color:black;font-size:16px;text-align:left;"> There is a `r paste0(any_atv_low_end_68, ' - ', any_atv_high_end_68)`% chance that an activation will occur in a given year.</p>
* <p style="color:black;font-size:16px;text-align:left;"> At least some activities are expected to be implemented ahead of `r paste0(any_det_low_end_68, ' - ', any_det_high_end_68)`% of shocks.</p>
* <p style="color:black;font-size:16px;text-align:left;"> The risk of a false alarm stands at `r paste0(any_far_low_end_68, ' - ', any_far_high_end_68)`% of activations.</p>
* <p style="color:black;font-size:16px;text-align:left;"> The mechanism is optimised to reduce the risk of failing to activate when there is a shock, at the cost of having false alarms. `r mitigation_warranted `</p>

&nbsp;

_________

## Trigger Mechanism Snapshot
<p style="color:black;font-size:16px;text-align:left;"> This table summarises the trigger mechanism and its performance in signaling severe shocks. It presents estimates of the likelihood that the threshold would be met in a given year as well as how often the trigger is expected to correctly recommend an activation or non-activation.</p> 

<p style="color:black;font-size:16px;text-align:left;"> The estimates are shown as ranges called confidence intervals that reflect the impact that data limitations (small or incomplete datasets, errors or imprecision in measurements, divergences between data sources) can have on the accuracy of performance estimates. Confidence intervals can inform decision-making by illustrating the most likely performance levels and by providing bounds for the plausible although unlikely extremes.</p> 

<br>

* <p style="color:black;font-size:16px;text-align:left;"> The <span style="font-weight: bold;">central value</span> is the most likely performance; </p>
* <p style="color:black;font-size:16px;text-align:left;"> Most often the trigger will perform within the <span style="font-weight: bold;">darker inner range</span> (68% of the time);</p>
* <p style="color:black;font-size:16px;text-align:left;"> There is high confidence that the trigger will perform within the <span style="font-weight: bold;">full coloured (grey or blue) range</span> (95% of the time).</p> 

<br>
<p style="color:black;font-size:16px;text-align:left;"> Confidence intervals in <span style="color: blue;font-weight: bold;">blue</span> represent activation likelihood, those in <span style="color: green;font-weight: bold;">green</span> represent desired outcomes (valid non-activations/activations) and those in <span style="color: red;font-weight: bold;"dou>red</span> represent undesired outcomes or errors.</p>

&nbsp;

```{r mechanism-snapshot, include = TRUE}

snapshot_df <- data.frame(row_name = c('Type', 
                                       'Monitored Area',
                                       'Activation Timepoints', 
                                       'Target Period', 
                                       'Lead Time', 
                                       'Data Source',
                                       'Historical Data Coverage',
                                       'Monitored by',
                                       'Activity Package', 
                                       'Funding Amount',
                                       'Probability of being met in a given year', 
                                       'Expected Cost',
                                       'Activations', 
                                       'Shocks'),
                       trigger_a = c(type_a, 
                                     monitored_area_a,
                                     timepoints_a, 
                                     targeted_window_a, 
                                     lead_time_a, 
                                     data_source_a,
                                     hist_period_a,
                                     monitored_by_a,
                                     package_a, 
                                     funding_a_formatted,
                                     "<img src=plots/trimmed_Trigger1_atv.png>",
                                     expected_cost_a,
                                     "<img src=plots/trimmed_Trigger1_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_Trigger1_det_tradeoff_bar.png>"),
                       trigger_b = c(type_b,
                                     monitored_area_b,
                                     timepoints_b, 
                                     targeted_window_b, 
                                     lead_time_b, 
                                     data_source_b,
                                     hist_period_b,
                                     monitored_by_b,
                                     package_b, 
                                     funding_b_formatted,
                                     "<img src=plots/trimmed_Trigger2_atv.png>", 
                                     expected_cost_b,
                                     "<img src=plots/trimmed_Trigger2_var_tradeoff_bar.png>",
                                     "<img src=plots/trimmed_Trigger2_det_tradeoff_bar.png>"),
                       trigger_c = c(type_c, 
                                     monitored_area_c,
                                     timepoints_c, 
                                     targeted_window_c, 
                                     lead_time_c, 
                                     data_source_c,
                                     hist_period_c,
                                     monitored_by_c,
                                     package_c, 
                                     funding_c_formatted,
                                    "<img src=plots/trimmed_Trigger3_atv.png>", 
                                    expected_cost_c,
                                     "<img src=plots/trimmed_Trigger3_var_tradeoff_bar.png>",
                                     "<img src=plots/trimmed_Trigger3_det_tradeoff_bar.png>"))

kbl(snapshot_df,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, # omit row numbers
    col.names = c("","Trigger 1", "Trigger 2", "Trigger 3"), 
    align = "rccc") %>%
  kable_minimal(c("hover", "striped")) %>%
  row_spec(c(1:12), extra_css = "line-height: 30px;") %>%
  row_spec(c(13:14), extra_css = "line-height: 80px;") %>%
  column_spec(column = 1, extra_css = "vertical-align: bottom;") %>%
  column_spec(column = c(2:4), width = "28em") %>%
  pack_rows("Description", start_row = 1, end_row = 8, color = '#1bb580') %>%
  pack_rows("Programming and Funding", start_row = 9, end_row = 12, color = '#1bb580') %>%
  pack_rows("Performance", start_row = 13, end_row = 14
            , color = '#1bb580') %>%
  kableExtra::footnote(general_title = "Notes",
                       general = trigger_table_footnote)

```

&nbsp;

## Activation Scenarios
<p style="color:black;font-size:16px;text-align:left;"> This section examines the likelihood of activation in a given year under various scenarios. The estimates are presented as confidence intervals to illustrate their inherent uncertainty. The graphs show how likely it is that a scenario will occur (<span style="color:  #007ce1;">darker inner band</span>, 68% of years) as well as the highest and lowest probability that a scenario is to occur (<span style="color:  `r alpha("#007ce1", 0.7)`;">full coloured band</span>, 95% of years).</p>

&nbsp;

```{r framework-activation-table}
activation_df <- data.frame(row_name = c(paste0('Disbursements', footnote_marker_number(1)), 
                                        paste0('Full', footnote_marker_number(2)),
                                        paste0('Package 1 only', footnote_marker_number(3)),
                                        paste0('Package 2 only', footnote_marker_number(4))),
                           atv_stats = c("<img src=plots/trimmed_framework-min_atv_tradeoff_bar.png>",  
                                         "<img src=plots/trimmed_framework-full_atv.png>",
                                         "<img src=plots/trimmed_framework-p1_atv.png>", 
                                         "<img src=plots/trimmed_framework-p2_atv.png>"))

kbl(activation_df,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, 
    col.names = c("",""), 
    align = "rl") %>%
  kable_styling(full_width = FALSE) %>%
  kable_minimal(c("hover")) %>%
  row_spec(c(1:4), extra_css = "line-height: 45px;") %>%
  column_spec(column = 1, width = "20em", extra_css = "vertical-align: bottom;") %>%
  column_spec(column = 2, width = "22em") %>%
  pack_rows("Activations", start_row = 1, end_row = 2, color = '#1bb580') %>%
  pack_rows("Disbursement Scenario Breakdown", 2, 4, color = 'black') %>%
  kableExtra::footnote(general_title = "Notes",
                       general = activation_table_footnote,
                       number = c("Any: At least one trigger is met. None: No trigger is met.",
                        full_atv_definition, 
                        package1_atv_definition,
                        package2_atv_definition))

```

&nbsp;

## Technical Team
<p style="color:black;font-size:16px;text-align:left;"> `r data_providers` have provided data and/or analytical support. A total of `r nbr_hist_shocks` historical occurrences of severe shocks were documented and used to develop the trigger mechanism. `r analysis` </p>

## Monitoring Process
<p style="color:black;font-size:16px;text-align:left;"> `r monitoring_process` </p>

## Learning Opportunities
<p style="color:black;font-size:16px;text-align:left;"> This table summarises the decisions, assumptions, and open questions that arose during trigger development. They are documented for transparency and to inform the learning agenda.</p>

&nbsp;

```{r learning-table}

learning_table <- read.csv(learning_table_filename)

kbl(learning_table,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, # omit row numbers
    col.names = c("Decision","Rationale", "Assumption", "Open Question"), 
    align = "llll") %>%
  kable_minimal(c("hover", "striped")) %>%
  column_spec(column = c(1:4), width = "28em")

```

&nbsp;

## Annex
<p style="color:black;font-size:16px;text-align:left;"> This table reports the performance metrics per activation timepoint for the triggers that have more than one opportunities for activation.</p>

&nbsp;

```{r timepoint-tables, include = TRUE}

timepoint_df1 <- data.frame(row_name = c('Triggering Prob.',
                                        'Activations', 
                                       'Shocks'),
                       trigger_jan = c("<img src=plots/trimmed_Jan_atv.png>",
                                     "<img src=plots/trimmed_Jan_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_Jan_det_tradeoff_bar.png>"),
                       trigger_feb = c("<img src=plots/trimmed_Feb_atv.png>",
                                     "<img src=plots/trimmed_Feb_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_Feb_det_tradeoff_bar.png>"),
                       trigger_mar = c("<img src=plots/trimmed_Mar_atv.png>", 
                                     "<img src=plots/trimmed_Mar_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_Mar_det_tradeoff_bar.png>"))

timepoint_df2 <- data.frame(row_name = c('Triggering Prob.',
                                        'Activations', 
                                       'Shocks'),
                       trigger_apr = c("<img src=plots/trimmed_Apr_atv.png>",
                                     "<img src=plots/trimmed_Apr_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_Apr_det_tradeoff_bar.png>"),
                       trigger_may = c("<img src=plots/trimmed_May_atv.png>",
                                     "<img src=plots/trimmed_May_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_May_det_tradeoff_bar.png>"),
                       trigger_jun = c("<img src=plots/trimmed_June_atv.png>",
                                     "<img src=plots/trimmed_June_var_tradeoff_bar.png>", 
                                     "<img src=plots/trimmed_June_det_tradeoff_bar.png>"))

kbl(timepoint_df1,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, 
    col.names = c("","January", "February", "March"), 
    align = "rccc") %>%
  kable_minimal(c("hover")) %>%
  row_spec(c(1:3), extra_css = "line-height: 60px;") %>%
  column_spec(column = 1, extra_css = "vertical-align: bottom", width = "10em")
```

&nbsp;

```{r timpoint_table2}
kbl(timepoint_df2,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, 
    col.names = c("","April", "May", "June"), 
    align = "rccc") %>%
  kable_minimal(c("hover")) %>%
  row_spec(c(1:3), extra_css = "line-height: 60px;") %>%
  column_spec(column = 1, extra_css = "vertical-align: bottom", width = "10em")
```

