---
title: '<img src="https://drive.google.com/uc?id=118y5T73-iSUZoAMtoJGddxq9QzD_GDKX" style="height:40px;float:left;margin:10px" /><img src="https://drive.google.com/uc?id=1fHQUzF3ZjaoHj9KQ33-94dK_X1hcmjzW" style="height:50px;float:left;margin:10px" />'
pagetitle: 'AA Trigger Report'
output:
  html_document:
    css: ../../country_template/docs/style.css
    includes:
      in_header: ../../country_template/docs/header.html
    toc: no
    toc_float: yes
    toc_depth: '4'
    df_print: paged
  html_notebook:
    css: ../../country_template/docs/style.css
    includes:
      in_header: ../../country_template/docs/header.html
    toc: no
    toc_float: yes
    toc_depth: '4'
    df_print: paged
editor_options:
  chunk_output_type: console
---
<style type="text/css">
.main-container {
  max-width: 90%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r README, echo = FALSE, message = FALSE, warning = FALSE}
### NOTE: The template is set up for two 3-timepoint predictive triggers and a 1-timepoint observational trigger
### NOTE: Performance metrics and their CIs should have been generated separately before the trigger report can be produced.

### Create a trigger_report folder under DATA_DIR > /private/exploration/{country_pcode3/trigger_performance}
### Copy the dummy_learning_table_for_template.csv file to the trigger_report folder on DATA_DIR; remove 'dummy" and "for_template" from filename; add the 3-letter pcode for the country at the very beginning of the filename (eg: ner_perf_metrics_table.csv)
### Create a "plots" folder sister to the RMD file in the country's `analyses` folder
### Update the "hard-coded-values" code chunk below
### If there are fewer or more than 3 triggers and/or fewer or more than 2 packages: Adjust the number of triggers and/or Packages throughout the Rmd (review the snapshot table and activation table carefully.) 
### Review, adjust or remove the country-specific code chunk
### Remove the Annex if no trigger has more than 1 activation timepoint or adjust the Annex as needed 
### Knit
```

```{r hard-coded-values, echo = FALSE, message = FALSE, warning = FALSE}
country_name <- "Chad"
country_pcode3 <- "tcd"
shock <- "Drought"
nbr_hist_shocks <- 5 # X is a number. #NER: Number of reported bad years 

# trigger A info  
name_a <- "Trigger1" # what name is the trigger listed as in the perf_metrics_table.csv files?
type_a <- 'Predictive' # Predictive or Observational?
monitored_area_a <- "Central Chad"
timepoints_a<- 'Mar, Apr' # eg: 'Jan, Feb, Mar'
package_a <- 'Package 1' # eg: 'Package 1'
funding_a <- 3.71 # input number only, no $ sign or 'm' for million. Formatting will be done automatically.
targeted_window_a <- 'Jul-Sep' # 3-letter names for first and last month. eg: 'Jul-Sep'. Forecasted period or period over which obs data are computed.
lead_time_a <- '4-3 months' # 'eg: 7-5 months', Change units as needed. Include all lead times if multiple activation timepoints
data_source_a <- 'IRI seasonal forecast' # name of product + provider
trigger_a <- '42.5% below average probability'
hist_period_a <- "2018 - present" # period for which historical data is available
monitored_by_a <- 'CHD' # owner of monitoring responsibilities

# trigger B info  
name_b <- "Trigger2"
type_b <- 'Predictive'
monitored_area_b <- "Central Chad"
timepoints_b <- 'May, Jun'
package_b <- 'Package 2'
funding_b <- 1.66
targeted_window_b <- 'Jul-Sep'
lead_time_b <- '2-1'
data_source_b <- 'IRI seasonal forecast'
trigger_b <- '42.5% below average probability'
hist_period_b <- "2018 - Present"
monitored_by_b <- 'CHD'
  
# trigger C info  
name_c <- "Trigger3"
type_c <- 'Observational'
monitored_area_c <- "Central Chad"
timepoints_c <- 'Sep'
package_c <- 'Package 2'
funding_c <- 4.63
targeted_window_c <- 'Apr-Sep'
lead_time_c <- 'N/A'
data_source_c <- 'Biomass'
trigger_c <- 'Seasonal anomaly below 80%'
hist_period_c <- "2000 - Present"
monitored_by_c <- 'CHD'

# Compile list of triggers to be included in the report
trigger_list <- c(name_c) # limited due to issues with historical data in Triggers 1 and 2
scenario_list <- c() # No scenarios given limits of data

# Table footnotes
trigger_table_footnote <- "Expected annual cost computed using most likely probability of activation (central value). All triggers can reach their threshold and activate independently from one another. Triggers 1 and 2 did not have sufficient data for testing performance on Chad, but likelihood of activation tested on global data. Trigger 3 performance assessed by comparing recommended activations with historical bad years (i.e., years with a shock). 'Bad years' were identified through convergence of evidence reviewed from historical CERF allocations, FEWS NET reports, EM-DAT data, World Bank statistics and the 2015 and 2017 multi-risk contingency plans for Chad."

full_atv_definition <- "A full activation occurs when all three triggers met and all packages are disbursed."
package1_atv_definition <- "Package 1 is activated when Trigger 1 is met."
package2_atv_definition <- "Package 2 is activated when Trigger 2 is met."
package3_atv_definition <- "Package 3 is activated when Trigger 3 is met."
activation_table_footnote <-  "" # additional notes, if needed

# Data providers
data_providers <- "IRI Columbia University, the Food and Agriculture Organization (FAO), and Action Against Hunger (ACF) Spain" # will be followed by "provide data and/or analytical support."

# Analysis
analysis <- "The triggers were developed by the Centre for Humanitarian Data close collaboration with OCHA, FAO, and other participating agencies."

# Monitoring process
monitoring_process <- "Between March and June inclusively, the Centre for Humanitarian Data will monitor IRI seasonal forecasts for Triggers 1 and 2. Seasonal forecasts are released on the 15th of each month, and if the 15th falls on a weekend, then it is released on the closest weekday. Within 24 hours of the update, the CHD team notifies the AA team, the Chad Humanitarian Coordinator, and the Chief of CERF by email whether or not the trigger is met. By September 1, updates to the Chad biomass data from ACF Spain should be published for the second dekad of August. Within 24 hours, the Centre for Humanitarian Data will calculate cumulative biomass, compare to the threshold, and report to the AA team, the Chad Humanitarian Coordinator, and the Chief of CERF if Trigger 3 has been met. In case of an activation, additional meteorological and agricultural data will be shared by the Centre for Humanitarian Data and/or FAO to inform response targeting."
  
```

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE) # do not print code by default
knitr::opts_chunk$set(include = TRUE) # include chunk output by default
knitr::opts_chunk$set(message = FALSE) # do not print messages by default
knitr::opts_chunk$set(warning = FALSE) # do not print warnings by default

options(scipen = 999) # turn off scientific notation
options(encoding = "UTF-8") # set encoding to UTF-8 instead of ANSI
```

```{r libraries, include = FALSE, warning = FALSE}
packages <- c('tidyverse', 'ggthemes', 'kableExtra', 'knitr', 'flextable')

# install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# load libraries 
lapply(packages, library, character.only = TRUE)

```


```{r paths}
data_folder_path <- paste0(Sys.getenv("AA_DATA_DIR"), "/private/exploration/", country_pcode3,"/trigger_performance/")
perf_metrics_95_filename <- paste0(data_folder_path, country_pcode3, "_perf_metrics_table_ci_0.95.csv")
perf_metrics_68_filename <- paste0(data_folder_path, country_pcode3, "_perf_metrics_table_ci_0.68.csv")
learning_table_filename <- paste0(data_folder_path, country_pcode3, "_learning_table.csv")
explanatory_ci_plot_path <- ("../../country_template/docs/explanatory_ci.png")
```

```{r functions}
source("../../country_template/docs/plotAtvCI.R")
source("../../country_template/docs/plotTradeoffCI.R")
```

```{r data-loading-and-formatting}

# read in performance metrics files
perf_metrics_95_data <- read.csv(perf_metrics_95_filename,
                              colClasses = c('character', 'character', 'numeric', 'character')
                              )

perf_metrics_68_data <- read.csv(perf_metrics_68_filename,
                              colClasses = c('character', 'character', 'numeric', 'character')
                              )

# format performance metrics values
perf_metrics_95_data$value <- round(100 * perf_metrics_95_data$value, 0)
perf_metrics_68_data$value <- round(100 * perf_metrics_68_data$value, 0)

# uniquely identify point labels
perf_metrics_95_data$upoint <- paste0(perf_metrics_95_data$point, "_95")
perf_metrics_68_data$upoint <- paste0(perf_metrics_68_data$point, "_68")

# combine datasets
perf_metrics_data <- rbind(perf_metrics_95_data, perf_metrics_68_data) %>%
                      select(-point) # remove non-unique point identifier

# format funding amounts
funding_a_formatted <- paste0('$',funding_a,'m')
funding_b_formatted <- paste0('$',funding_b,'m')
funding_c_formatted <- paste0('$',funding_c,'m')

# expected cost amounts (probability of activation multiplied by funding amount)
expected_cost_a <- paste0('$', round(funding_a * (perf_metrics_data %>% filter(trigger == name_a & metric == "atv" & upoint == 'central_95') %>% select(value) %>% as.numeric()/100), 2), 'm')
expected_cost_b <- paste0('$', round(funding_b * (perf_metrics_data %>% filter(trigger == name_b & metric == "atv" & upoint == 'central_95') %>% select(value) %>% as.numeric()/100), 2), 'm')
expected_cost_c <- paste0('$', round(funding_c * (perf_metrics_data %>% filter(trigger == name_c & metric == "atv" & upoint == 'central_95') %>% select(value) %>% as.numeric()/100), 2), 'm')

# probabilities just for triggers 1 and 2 that are global
prob_a <- perf_metrics_95_data %>% filter(trigger == name_a & metric == "atv") %>% pull(value)
prob_b <- perf_metrics_95_data %>% filter(trigger == name_b & metric == "atv") %>% pull(value)
```

```{r atv-CIs}

# Create a CI visual for atv for each trigger, activation timepoint, and activation scenario beside None and Any

for(trigger in c(trigger_list, scenario_list)) {

  # produce segment_dimensions and plots for triggers 
  output <- plotAtvCI(trigger_id = trigger)
  
  # save plot as png
  filename <- paste0(trigger, "_atv.png")
  png(filename = paste0("plots/", filename), width = 500, height = 400, units = "px")
  print(output$p)
  invisible(dev.off())
}

```

```{r tradeoff-CIs}

# Tradeoff CIs for individual triggers

for(trigger in c(trigger_list)) {
  for (metric in c('var', 'det')) {
  
  output <- plotTradeoffCI(trigger_id = trigger, 
                   left_metric_name = metric)
  
  # save plot as png
  filename <- paste0(trigger, "_", metric, "_tradeoff_bar.png")
  png(filename = paste0("plots/", filename), width = 500, height = 400, units = "px")
  print(output$p)
  invisible(dev.off())
  }
}

```

```{r file-formatting-and-saving}

# list plots to be trimmed
plots_to_trim <- grep(list.files(path = "./plots"), 
                      pattern = 'trimmed_', 
                      invert = TRUE, 
                      value = TRUE) # excludes 'trimmed_' graphs, which will be overwritten

# trim plots
for (filename in plots_to_trim) {

  # read in the image
  original_plot <- magick::image_read(paste0("plots/", filename))
  
  # trim the image
  trimmed_plot <- magick::image_trim(original_plot)
  
  # save trimmed plot
  magick::image_write(trimmed_plot, path = paste0("plots/trimmed_", filename), format = "png")

}

```

```{r key-messages-metrics}
t3_atv_low_end_68 <- perf_metrics_data %>% filter(trigger == 'Trigger3' & metric == 'atv') %>% filter(upoint == 'low_end_68') %>% select(value) %>% as.numeric()

t3_atv_high_end_68 <- perf_metrics_data %>% filter(trigger == 'Trigger3' & metric == 'atv') %>% filter(upoint == 'high_end_68') %>% select(value) %>% as.numeric()

t12_atv_central <- perf_metrics_data %>% filter(trigger == 'Trigger1-2' & metric == 'atv') %>% filter(upoint == 'central_95') %>% select(value) %>% as.numeric()
t3_det_low_end_68 <- perf_metrics_data %>% filter(trigger == 'Trigger3' & metric == 'det') %>% filter(upoint == 'low_end_68') %>% select(value) %>% as.numeric()

t3_det_high_end_68 <- perf_metrics_data %>% filter(trigger == 'Trigger3' & metric == 'det') %>% filter(upoint == 'high_end_68') %>% select(value) %>% as.numeric()

t3_far_low_end_68 <- perf_metrics_data %>% filter(trigger == 'Trigger3' & metric == 'far') %>% filter(upoint == 'low_end_68') %>% select(value) %>% as.numeric()

t3_far_high_end_68 <- perf_metrics_data %>% filter(trigger == 'Trigger3' & metric == 'far') %>% filter(upoint == 'high_end_68') %>% select(value) %>% as.numeric()

mitigation_warranted <- ifelse(t3_far_low_end_68 >= 40, "Mitigation is recommended to minimise the negative impact of a false alarm due to its moderate to high likelihood.", "")

```

# Anticipatory Action Trigger Mechanism Report
<p style="color:#007ce1;font-size:22px;text-align:center;"> `r paste0(shock, ' in ', country_name)`</p> 
<p style="color:black;font-size:14px;text-align:right;"> Last updated: `r Sys.Date()`</p> 

## Key Messages
* <span style="color: blue;">LIMITED HISTORICAL DATA</span> <p style="color:black;font-size:16px;text-align:left;"> Triggers 1 and 2 rely on the IRI seasonal forecast, with limited data from 2018 to today. Likelihood of activation estimated based on global data, for Triggers 1 and 2 together, but not together with Trigger 3, and no performance metrics are calculated.</p>
* <span style="color: blue;">ACTIVATION</span> <p style="color:black;font-size:16px;text-align:left;"> There is a `r paste0(t3_atv_low_end_68, ' - ', t3_atv_high_end_68)`% chance that Trigger 3 will activate in a given year. Globally, thresholds are met for Trigger 1 or 2 `r t12_atv_central`% of the time, however there is high uncertainty on how frequently this would occur in `r country_name`.</p>
* <span style="color: blue;">SHOCK DETECTION</span> <p style="color:black;font-size:16px;text-align:left;"> Package 3 is expected to be implemented ahead of `r paste0(t3_det_low_end_68, ' - ', t3_det_high_end_68)`% of shocks.</p>
* <span style="color: blue;">FALSE ALARMS</span> <p style="color:black;font-size:16px;text-align:left;"> The risk of activating Package 3 in the absence of a severe shock stands at `r paste0(t3_far_low_end_68, ' - ', t3_far_high_end_68)`% of activations.</p>
* <span style="color: blue;">NO REGRETS BIAS</span> <p style="color:black;font-size:16px;text-align:left;"> The mechanism is optimised to reduce the risk of failing to activate when there is a shock, at the cost of having false alarms. `r mitigation_warranted `</p>

&nbsp;

_________

## About Metric Estimates

<p style="color:black;font-size:16px;text-align:left;"> The estimates in this report are shown as ranges called confidence intervals. The confidence intervals reflect the probabilistic nature of the estimates as well as the impact of data limitations (small or incomplete datasets, errors or imprecision in measurements, divergences between data sources, etc.) on the accuracy of estimates. They can inform decision-making by illustrating the most likely performance levels and by providing bounds for the possible albeit unlikely extremes.</p> 

<br>

```{r explanatory_ci, echo=FALSE, out.width= "40%", out.extra='style="float:right; padding:10px"'}
explanatory_ci <- magick::image_read(explanatory_ci_plot_path)

explanatory_ci
```

* <p style="color:black;font-size:16px;text-align:left;"> The <span style="font-weight: bold;">central value</span> is the most likely performance; </p>
* <p style="color:black;font-size:16px;text-align:left;"> Most often the trigger will perform within the <span style="font-weight: bold;">darker inner range</span> (68% of the time);</p>
* <p style="color:black;font-size:16px;text-align:left;"> There is high confidence that the trigger will perform within the <span style="font-weight: bold;">full coloured (grey or blue) range</span> (95% of the time).</p> 

<br>

## Colour Coding
<p style="color:black;font-size:16px;text-align:left;"> In the visualizations <span style="color: blue;font-weight: bold;">blue</span> represents activation likelihood, <span style="color: green;font-weight: bold;">green</span> represents desired outcomes (valid non-activations/activations) and <span style="color: red;font-weight: bold;"dou>red</span> represents undesired outcomes or errors.</p>

&nbsp;

## Trigger Mechanism Snapshot
<p style="color:black;font-size:16px;text-align:left;"> This table summarises the trigger mechanism and its performance in signaling severe shocks. It presents estimates of the likelihood that the threshold would be met in a given year as well as how often the trigger is expected to correctly recommend an activation or non-activation.</p> 

&nbsp;
```{r mechanism-snapshot, include = TRUE}

snapshot_df <- data.frame(row_name = c('Type', 
                                       'Monitored Area',
                                       'Activation Timepoints', 
                                       'Target Period', 
                                       'Lead Time', 
                                       'Data Source',
                                       'Trigger',
                                       'Historical Data Coverage',
                                       'Monitored by',
                                       'Activity Package', 
                                       'Probability of being met in a given year', 
                                       'Funding Amount',
                                       'Expected Annual Cost',
                                       'All Activations (%)', 
                                       'All Shocks (%)'),
                       trigger_a = c(type_a, 
                                     monitored_area_a,
                                     timepoints_a, 
                                     targeted_window_a, 
                                     lead_time_a, 
                                     data_source_a,
                                     trigger_a,
                                     hist_period_a,
                                     monitored_by_a,
                                     package_a, 
                                     paste0(prob_a, "%, based on global IRI data, but very high uncertainty"),
                                     funding_a_formatted,
                                     expected_cost_a,
                                     "N/A", 
                                     "N/A"),
                       trigger_b = c(type_b,
                                     monitored_area_b,
                                     timepoints_b, 
                                     targeted_window_b, 
                                     lead_time_b, 
                                     data_source_b,
                                     trigger_b,
                                     hist_period_b,
                                     monitored_by_b,
                                     package_b, 
                                     paste0(prob_b, "%, based on global IRI data, but very high uncertainty"), 
                                     funding_b_formatted,
                                     expected_cost_b,
                                     "N/A",
                                     "N/A"),
                       trigger_c = c(type_c, 
                                     monitored_area_c,
                                     timepoints_c, 
                                     targeted_window_c, 
                                     lead_time_c, 
                                     data_source_c,
                                     trigger_c,
                                     hist_period_c,
                                     monitored_by_c,
                                     package_c, 
                                    "<img src=plots/trimmed_Trigger3_atv.png>", 
                                     funding_c_formatted,
                                     expected_cost_c,
                                     "<img src=plots/trimmed_Trigger3_var_tradeoff_bar.png>",
                                     "<img src=plots/trimmed_Trigger3_det_tradeoff_bar.png>"))

kbl(snapshot_df,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, # omit row numbers
    col.names = c("","Trigger 1", "Trigger 2", "Trigger 3"), 
    align = "rccc") %>%
  kable_minimal(c("hover", "striped")) %>%
  row_spec(1:15, extra_css = "line-height: 30px;") %>%
  column_spec(column = 1, extra_css = "vertical-align: middle;") %>%
  column_spec(column = c(2:4), width = "28em") %>%
  pack_rows("Description", start_row = 1, end_row = 8, color = '#1bb580') %>%
  pack_rows("Programming and Funding", start_row = 9, end_row = 12, color = '#1bb580') %>%
  pack_rows("Performance", start_row = 14, end_row = 15
            , color = '#1bb580') %>%
  kableExtra::footnote(general_title = "Notes",
                       general = trigger_table_footnote)

```

&nbsp;

## Activation Scenarios
<p style="color:black;font-size:16px;text-align:left;"> Due to the lack of historical data for the IRI seasonal forecast, we do not assess the likelihood of activation scenarios together. Refer to the likelihood of the individual triggers activating.</p>

&nbsp;

## Technical Team
<p style="color:black;font-size:16px;text-align:left;"> `r data_providers` have provided data and/or analytical support. A total of `r nbr_hist_shocks` historical occurrences of severe shocks were documented and used to develop the trigger mechanism. `r analysis` </p>

## Monitoring Process
<p style="color:black;font-size:16px;text-align:left;"> `r monitoring_process` </p>

## Learning Opportunities
<p style="color:black;font-size:16px;text-align:left;"> This table summarises the decisions, assumptions, and open questions that arose during trigger development. They are documented for transparency and to inform the learning agenda.</p>

&nbsp;

```{r learning-table}

learning_table <- read.csv(learning_table_filename)

kbl(learning_table,
    format = 'html',
    escape = FALSE,
    row.names = FALSE, # omit row numbers
    col.names = c("Decision","Rationale", "Assumption", "Open Question"), 
    align = "llll") %>%
  kable_minimal(c("hover", "striped")) %>%
  column_spec(column = c(1:4), width = "28em")

```

&nbsp;
