---
title: "MWI: Compare ARC2 and CHIRPS"
author: "Seth Caldwell"
date: "13/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(fuzzyjoin)
library(lubridate)
library(sf)
library(cowplot)

# file paths

data_dir <- Sys.getenv("AA_DATA_DIR")
expl_dir <- paste0(data_dir, "/public/exploration/mwi")
process_dir <- paste0(data_dir, "/public/processed/mwi")
shapefile_dir <- paste0(data_dir, "/public/raw/mwi/cod_ab/mwi_adm_nso_20181016_shp")
```

## Initial data loading and wrangling

Alright, so first we need to load in the data that has already been processed. We have
3 files, one for CHIRPS data, and 2 for ARC2 data, all aggregated at the ADM2 level.
All datasets contain dry spells identified through that particular dataset, and filtered
to only occur during rainy seasons identified using CHIRPS data.

Since the rainy season definitions would slightly vary using ARC2 definitions,
some of the dry spells recorded captured in those datasets may be instead the
onset of the dry season and not captured if we defined the rainy season using ARC2
data. After some exploration, I found that for ARC2 dry spells with a duration `> 30`,
for both the centroids and touching method, only a single one of the
20 and 18 dry spells respectively overlapped with a CHIRPS identified dry spell, and
none had a dry spell confirmation in January or February. Thus, I explicitly filter
these out prior to comparing the various dry spell identification methods.

```{r loading}
# data loading

arc2_center_df <- read_csv(paste0(expl_dir, "/arc2/mwi_arc2_centroid_dry_spells_during_rainy_season.csv"),
                           col_types = "_ccDDDdd") %>%
  select(pcode,
         ADM2_EN,
         arc2_center_start = dry_spell_first_date,
         arc2_center_confirmation = dry_spell_confirmation,
         arc2_center_end = dry_spell_last_date,
         arc2_center_duration = dry_spell_duration) %>%
  filter(arc2_center_duration <= 30)

arc2_touch_df <- read_csv(paste0(expl_dir, "/arc2/mwi_arc2_touching_dry_spells_during_rainy_season.csv"),
                          col_types = "_ccDDDd") %>%
  select(pcode,
         ADM2_EN,
         arc2_touch_start = dry_spell_first_date,
         arc2_touch_confirmation = dry_spell_confirmation,
         arc2_touch_end = dry_spell_last_date,
         arc2_touch_duration = dry_spell_duration) %>%
  filter(arc2_touch_duration <= 30)

chirps_df <- read_csv(paste0(
  process_dir, 
  "/dry_spells/v1/archive/dry_spells_during_rainy_season_list_2000_2020_mean_back.csv"
)) %>%
  transmute(pcode,
            ADM2_EN,
            chirps_start = dry_spell_first_date,
            chirps_confirmation = dry_spell_first_date + 13,
            chirps_end = dry_spell_last_date,
            chirps_duration = dry_spell_duration)

# shapefile loading
adm0 <- st_read(paste0(shapefile_dir, "/mwi_admbnda_adm0_nso_20181016.shp"))
adm1 <- st_read(paste0(shapefile_dir, "/mwi_admbnda_adm1_nso_20181016.shp"))
adm2 <- st_read(paste0(shapefile_dir, "/mwi_admbnda_adm2_nso_20181016.shp"))

# daily values loading
arc2_daily_df <- read_csv(paste0(expl_dir, "/arc2/mwi_arc2_precip_long_raw.csv"),
                          col_types = "____c_______D__dd______") %>%
  rename(pcode = ADM2_PCODE,
         arc2_centroid = mean_cell,
         arc2_touch = mean_cell_touched)

chirps_daily_df <- readRDS(paste0(process_dir, "/dry_spells/v1/data_long_mean_values_2000_2021.RDS")) %>%
  ungroup() %>%
  select(pcode, date, chirps = total_prec)

daily_df <- inner_join(arc2_daily_df,
                       chirps_daily_df,
                       by = c("pcode", "date"))
```

With the 3 separate datasets loaded, we can use the `fuzzyjoin` package to join the
datasets together (with some repititive filtering after) to capture all overlapping
dry spells to examine in detail how the 3 separate methods for dry spell identification
compare for each incident in particular.

```{r wrangling}
# joining ARC2 touch and center dry spells to get all overlaps

arc2_center_lj1 <- fuzzy_full_join(
  arc2_center_df,
  arc2_touch_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_start" = "arc2_touch_start",
    "arc2_center_start" = "arc2_touch_end"
  ))

arc2_center_lj2 <- fuzzy_full_join(
  arc2_center_df,
  arc2_touch_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_end" = "arc2_touch_start",
    "arc2_center_end" = "arc2_touch_end"
  ))

arc2_overlap_df <- bind_rows(arc2_center_lj1,
          arc2_center_lj2) %>%
  mutate(pcode.x = ifelse(is.na(pcode.x), pcode.y, pcode.x)) %>%
  rename(pcode = pcode.x) %>%
  select(-pcode.y) %>%
  distinct() %>%
  group_by(pcode, arc2_center_start) %>%
  filter(!(n() > 1 & is.na(arc2_touch_start))) %>% # drop duplicate row of center
  ungroup()

# now joining with CHIRPS to get additional overlaps

arc2_chirps_lj1 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_start" = "chirps_start",
    "arc2_center_start" = "chirps_end"
  ))

arc2_chirps_lj2 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_end" = "chirps_start",
    "arc2_center_end" = "chirps_end"
  ))

arc2_chirps_lj3 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_touch_start" = "chirps_start",
    "arc2_touch_start" = "chirps_end"
  ))

arc2_chirps_lj4 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_touch_end" = "chirps_start",
    "arc2_touch_end" = "chirps_end"
  ))

arc2_chirps_df <- bind_rows(arc2_chirps_lj1,
                            arc2_chirps_lj2,
                            arc2_chirps_lj3,
                            arc2_chirps_lj4) %>%
  distinct() %>%
  mutate(pcode.x = ifelse(is.na(pcode.x), pcode.y, pcode.x)) %>%
  rename(pcode = pcode.x) %>%
  select(-pcode.y) %>% 
  # clean up duplicate joined rows
  group_by(pcode, arc2_center_start, arc2_touch_start) %>%
  filter(!(n() > 1 & is.na(chirps_start))) %>%
  group_by(pcode, arc2_center_start, chirps_start) %>%
  filter(!(n() > 1 & is.na(arc2_touch_start))) %>%
  group_by(pcode, chirps_start, arc2_touch_start) %>%
  filter(!(n() > 1 & is.na(arc2_center_start))) %>%
  group_by(pcode, chirps_start) %>%
  filter(!(n() > 1 & is.na(arc2_center_start) & !is.na(chirps_start))) %>%
  ungroup() %>%
  select(-starts_with("ADM2_EN"))

# get simple ADM2_EN data set
adm2_df <- map(list(
  arc2_center_df,
  arc2_touch_df,
  chirps_df
), ~ select(.x, pcode, ADM2_EN)) %>%
  bind_rows() %>%
  distinct()

# add ADM2_EN to main data frame
arc2_chirps_df <- arc2_chirps_df %>%
  left_join(adm2_df, by = "pcode")
```

### Simple comparison

At a basic level, we detect about 1/3 more dry spells using ARC2 data than using CHIRPS data,
with `r nrow(arc2_center_df)` and `r nrow(arc2_touch_df)` detected with the centroid and touching methods
respectively, and only `r nrow(chirps_df)` using CHIRPS. Let's compare how and when
dry spells are detected by each method.

```{r detection-matrix}
arc2_chirps_df <- mutate(arc2_chirps_df,
  chirps_ds = !is.na(chirps_confirmation),
  arc2_centroid_ds = !is.na(arc2_center_confirmation),
  arc2_touch_ds = !is.na(arc2_touch_confirmation)
)

arc2_chirps_df %>%
  select(CHIRPS = chirps_ds,
         `ARC2 (Centroid)` = arc2_centroid_ds,
         `ARC2 (Touching)` = arc2_touch_ds) %>%
  group_by_all() %>%
  summarize(Occurrences = n(),
            .groups = "drop") %>%
  mutate(across(-Occurrences, ~ ifelse(.x, "Detected", "Not detected"))) %>%
  arrange(desc(Occurrences)) %>%
  knitr::kable()
```

We can see above a breakdown of dry spell detection by method. Interestingly, where
CHIRPS detects a dry spells, both ARC2 methods either also detect one or don't.

```{r onset-comparison, echo = FALSE}
arc2_chirps_df %>%
  filter(!is.na(chirps_start), !is.na(arc2_center_start)) %>%
  mutate(center_chirps = chirps_start - arc2_center_start,
         touch_chirps = chirps_start - arc2_touch_start) %>%
  arrange(touch_chirps) %>%
  mutate(row_id = row_number()) %>%
  pivot_longer(c(center_chirps, touch_chirps),
               names_to = "arc2_method",
               values_to = "day_diff") %>%
  mutate(arc2_method = ifelse(arc2_method == "touch_chirps",
                              "Centroid / Touching",
                              "Centroid only")) %>%
  ggplot(aes(x = row_id, y = day_diff, color = arc2_method)) +
  geom_point(position = "dodge", size = 3) +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  labs(y = "Days since ARC2 detection",
       title = "CHIRPS dry spell detection relative to ARC2 dry spell, days between") +
  scale_color_manual(values = c("Centroid / Touching" = "black", "Centroid only" = "grey"))
```

From above, we can see that the dry spell detection using CHIRPS always detects the dry
spell **after** it has been detected by ARC2. This is reassuring news for the switch
to ARC2 for the observational trigger. However, what does it look like where
CHIRPS has data and ARC2 is missing it?

### Geographic comparison

```{r geo-compare}
arc2_missing_plot <- arc2_chirps_df %>%
  filter(chirps_ds, !arc2_touch_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using CHIRPS") +
  theme(plot.title.position = "plot")

chirps_missing_plot <- arc2_chirps_df %>%
  filter(!chirps_ds, arc2_touch_ds, arc2_centroid_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using ARC2",
       subtitle = "(both centroid and touching method)") +
  theme(plot.title.position = "plot")

plot_grid(arc2_missing_plot, chirps_missing_plot)
```

```{r geo-compare-arc2}
centroid_missing_plot <- arc2_chirps_df %>%
  filter(arc2_centroid_ds, !arc2_touch_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using centroids") +
  theme(plot.title.position = "plot")

touch_missing_plot <- arc2_chirps_df %>%
  filter(!arc2_centroid_ds, arc2_touch_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using touching") +
  theme(plot.title.position = "plot")

plot_grid(centroid_missing_plot, touch_missing_plot)
```

### Trigger activation

Let's consider if we're using these datasets for trigger activation and how they would
compare. We will only consider dry spells that are confirmed at some point in January
and February. Let's just quickly look at the number of dry spells detected this time period
for each method to start.

```{r trigger-chirps}
chirps_df %>%
  filter(month(chirps_confirmation) <= 2) %>%
  mutate(trigger_year = year(chirps_confirmation)) %>%
  group_by(trigger_year) %>%
  summarize(n_triggers = n(),
            .groups = "drop")
```

```{r trigger-center}
arc2_center_df %>%
  filter(month(arc2_center_confirmation) <= 2) %>%
  mutate(trigger_year = year(arc2_center_confirmation)) %>%
  group_by(trigger_year) %>%
  summarize(n_triggers = n(),
            .groups = "drop")
```

```{r trigger-touch}
arc2_touch_df %>%
  filter(month(arc2_touch_confirmation) <= 2) %>%
  mutate(trigger_year = year(arc2_touch_confirmation)) %>%
  group_by(trigger_year) %>%
  summarize(n_triggers = n(),
            .groups = "drop")
```

Given that the 2 ARC2 methods are so similar, we likely only need to look at the centroid method.
This is because based on the matrix above, there are only 6 dry spells detected through
the touch method not detected through the centroid method, but all but one of those
is a city area. Alternatively, the centroid method would result in 8 triggers not captured
by the touch method. However, only one of these is an urban area and the other 7
are south-central districts in the southern region.

```{r get-rid-of-touch}
arc2_chirps_df %>%
  filter(arc2_touch_ds, !arc2_centroid_ds) %>%
  transmute(ADM2_EN, trigger_year = year(arc2_touch_confirmation))
```

```{r keep-that-centroid}
arc2_chirps_df %>%
  filter(!arc2_touch_ds, arc2_centroid_ds) %>%
  transmute(ADM2_EN, trigger_year = year(arc2_center_confirmation))
```

Although the centroid method fails to capture Zomba City (due to a lack of raster
centroid for ARC2 data), it seems sufficient to consider vs. CHIRPS excluding the
touching raster method.

### Daily value comparisons

First, let's do some 2d histograms to examine the distribution of various calculation methods.
What do we see between the centroid and touching raster methods applied?

```{r 2d-histo-arc}
daily_df %>%
  ggplot(aes(x = arc2_centroid,
             y = arc2_touch)) +
  geom_bin2d(bins = 200) +
  theme_bw() +
  geom_abline(intercept = 0)
```

While it looks like the touching method might skew higher (if just slightly) at
higher levels of rainfall, we are more concerned at their behavior at lower levels.

```{r 2d-histo-arc-1}
daily_df %>%
  filter(arc2_centroid <= 1 | arc2_touch <= 1) %>%
  ggplot(aes(x = arc2_centroid,
             y = arc2_touch)) +
  geom_bin2d(bins = 50) +
  theme_bw() +
  geom_abline(intercept = 0)
```

And above we can see that the centroid method in general trends much higher at these levels.
This intuitively makes sense as the touching method takes the average of more raster
cells than the centroid method, so will tend to regress toward the mean at the extremes.
Anyway, given their similar results in identifying dry spells and the results above,
should be fine to simply compare the centroid method against CHIRPS and not the
touching method.

```{r 2d-histo-chirps}
daily_df %>%
  ggplot(aes(x = arc2_centroid,
             y = chirps)) +
  geom_bin2d(bins = 200) +
  theme_bw() +
  geom_abline(intercept = 0)
```

```{r 2d-histo-chirps-2}
daily_df %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2)%>%
  ggplot(aes(x = arc2_centroid,
             y = chirps)) +
  geom_bin2d(bins = 200) +
  theme_bw() +
  geom_abline(intercept = 0)
```

The above graph is obviously difficult to interpret given some of the extreme values.
We could limit the axes, but might be more intuititive to graph instead the differences.
```{r diff-chirps}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw()
```

```{r diff-chirps-2}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw()
```

```{r diff-chirps-excl0}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 6000))
```

