---
title: "MWI: Compare ARC2 and CHIRPS"
author: "Seth Caldwell"
date: "13/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(fuzzyjoin)
library(lubridate)
library(sf)
library(cowplot)

# file paths

data_dir <- Sys.getenv("AA_DATA_DIR")
expl_dir <- paste0(data_dir, "/public/exploration/mwi")
process_dir <- paste0(data_dir, "/public/processed/mwi")
shapefile_dir <- paste0(data_dir, "/public/raw/mwi/cod_ab/mwi_adm_nso_20181016_shp")
```

## Initial data loading and wrangling

Alright, so first we need to load in the data that has already been processed. We have
3 files, one for CHIRPS data, and 2 for ARC2 data, all aggregated at the ADM2 level.
All datasets contain dry spells identified through that particular dataset, and filtered
to only occur during rainy seasons identified using CHIRPS data.

Since the rainy season definitions would slightly vary using ARC2 definitions,
some of the dry spells recorded captured in those datasets may be instead the
onset of the dry season and not captured if we defined the rainy season using ARC2
data. After some exploration, I found that for ARC2 dry spells with a duration `> 30`,
for both the centroids and touching method, only a single one of those
20 and 18 dry spells respectively overlapped with a CHIRPS identified dry spell, and
none had a dry spell confirmation in January or February. Thus, I explicitly filter
these 38 dry spells out from the two methods prior to comparing the various dry
spell identification methods.

```{r loading}
# data loading

arc2_center_df <- read_csv(paste0(expl_dir, "/arc2/mwi_arc2_centroid_dry_spells_during_rainy_season.csv"),
                           col_types = "_ccDDDdd") %>%
  select(pcode,
         ADM2_EN,
         arc2_center_start = dry_spell_first_date,
         arc2_center_confirmation = dry_spell_confirmation,
         arc2_center_end = dry_spell_last_date,
         arc2_center_duration = dry_spell_duration) %>%
  filter(arc2_center_duration <= 30)

arc2_touch_df <- read_csv(paste0(expl_dir, "/arc2/mwi_arc2_touching_dry_spells_during_rainy_season.csv"),
                          col_types = "_ccDDDd") %>%
  select(pcode,
         ADM2_EN,
         arc2_touch_start = dry_spell_first_date,
         arc2_touch_confirmation = dry_spell_confirmation,
         arc2_touch_end = dry_spell_last_date,
         arc2_touch_duration = dry_spell_duration) %>%
  filter(arc2_touch_duration <= 30)

chirps_df <- read_csv(paste0(
  process_dir, 
  "/dry_spells/v1/archive/dry_spells_during_rainy_season_list_2000_2020_mean_back.csv"
)) %>%
  transmute(pcode,
            ADM2_EN,
            chirps_start = dry_spell_first_date,
            chirps_confirmation = dry_spell_first_date + 13,
            chirps_end = dry_spell_last_date,
            chirps_duration = dry_spell_duration) %>%
  filter(str_detect(ADM2_EN, "MW3"))

# shapefile loading
adm0 <- st_read(paste0(shapefile_dir, "/mwi_admbnda_adm0_nso_20181016.shp"))
adm1 <- st_read(paste0(shapefile_dir, "/mwi_admbnda_adm1_nso_20181016.shp"))
adm2 <- st_read(paste0(shapefile_dir, "/mwi_admbnda_adm2_nso_20181016.shp"))

# daily values loading
arc2_daily_df <- read_csv(paste0(expl_dir, "/arc2/mwi_arc2_precip_long_raw.csv"),
                          col_types = "____c_______D__dd______") %>%
  rename(pcode = ADM2_PCODE,
         arc2_centroid = mean_cell,
         arc2_touch = mean_cell_touched)

chirps_daily_df <- readRDS(paste0(process_dir, "/dry_spells/v1/data_long_mean_values_2000_2021.RDS")) %>%
  ungroup() %>%
  select(pcode, date, chirps = total_prec)

daily_df <- inner_join(arc2_daily_df,
                       chirps_daily_df,
                       by = c("pcode", "date"))
```

With the 3 separate datasets loaded, we can use the `fuzzyjoin` package to join the
datasets together (with some repititive filtering after) to capture all overlapping
dry spells to examine in detail how the 3 separate methods for dry spell identification
compare for each incident in particular.

```{r wrangling}
# joining ARC2 touch and center dry spells to get all overlaps

arc2_center_lj1 <- fuzzy_full_join(
  arc2_center_df,
  arc2_touch_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_start" = "arc2_touch_start",
    "arc2_center_start" = "arc2_touch_end"
  ))

arc2_center_lj2 <- fuzzy_full_join(
  arc2_center_df,
  arc2_touch_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_end" = "arc2_touch_start",
    "arc2_center_end" = "arc2_touch_end"
  ))

arc2_center_lj3 <- fuzzy_full_join(
  arc2_touch_df,
  arc2_center_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_touch_end" = "arc2_center_start",
    "arc2_touch_end" = "arc2_center_end"
  ))

arc2_center_lj4 <- fuzzy_full_join(
  arc2_touch_df,
  arc2_center_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_touch_start" = "arc2_center_start",
    "arc2_touch_start" = "arc2_center_end"
  ))

arc2_overlap_df <- bind_rows(arc2_center_lj1,
                             arc2_center_lj2,
                             arc2_center_lj3,
                             arc2_center_lj4) %>%
  mutate(pcode.x = ifelse(is.na(pcode.x), pcode.y, pcode.x)) %>%
  rename(pcode = pcode.x) %>%
  select(-pcode.y) %>%
  distinct() %>%
  group_by(pcode, arc2_center_start) %>%
  filter(!(n() > 1 & is.na(arc2_touch_start))) %>% # drop duplicate row of center
  ungroup()

# now joining with CHIRPS to get additional overlaps

arc2_chirps_lj1 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_start" = "chirps_start",
    "arc2_center_start" = "chirps_end"
  ))

arc2_chirps_lj2 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_center_end" = "chirps_start",
    "arc2_center_end" = "chirps_end"
  ))

arc2_chirps_lj3 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_touch_start" = "chirps_start",
    "arc2_touch_start" = "chirps_end"
  ))

arc2_chirps_lj4 <- fuzzy_full_join(
  arc2_overlap_df,
  chirps_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "arc2_touch_end" = "chirps_start",
    "arc2_touch_end" = "chirps_end"
  ))

arc2_chirps_lj5 <- fuzzy_full_join(
  chirps_df,
  arc2_overlap_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "chirps_start" = "arc2_touch_start",
    "chirps_start" = "arc2_touch_end"
  ))

arc2_chirps_lj6 <- fuzzy_full_join(
  chirps_df,
  arc2_overlap_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "chirps_end" = "arc2_touch_start",
    "chirps_end" = "arc2_touch_end"
  ))

arc2_chirps_lj7 <- fuzzy_full_join(
  chirps_df,
  arc2_overlap_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "chirps_start" = "arc2_center_start",
    "chirps_start" = "arc2_center_end"
  ))

arc2_chirps_lj8 <- fuzzy_full_join(
  chirps_df,
  arc2_overlap_df,
  match_fun = list(`==`, `>=`, `<=`),
  by = c(
    "pcode" = "pcode",
    "chirps_end" = "arc2_center_start",
    "chirps_end" = "arc2_center_end"
  ))

arc2_chirps_df <- bind_rows(arc2_chirps_lj1,
                            arc2_chirps_lj2,
                            arc2_chirps_lj3,
                            arc2_chirps_lj4,
                            arc2_chirps_lj5,
                            arc2_chirps_lj6,
                            arc2_chirps_lj7,
                            arc2_chirps_lj8) %>%
  mutate(pcode.x = ifelse(is.na(pcode.x), pcode.y, pcode.x)) %>%
  rename(pcode = pcode.x) %>%
  select(-pcode.y, -starts_with("ADM2_EN")) %>% 
  distinct() %>%
  # clean up duplicate joined rows
  group_by(pcode, arc2_center_start, arc2_touch_start) %>%
  filter(!(n() > 1 & is.na(chirps_start) & !is.na(arc2_center_start) & !is.na(arc2_touch_start))) %>%
  group_by(pcode, arc2_center_start, chirps_start) %>%
  filter(!(n() > 1 & is.na(arc2_touch_start) & !is.na(arc2_center_start) & !is.na(chirps_start))) %>%
  group_by(pcode, chirps_start, arc2_touch_start) %>%
  filter(!(n() > 1 & is.na(arc2_center_start) & !is.na(chirps_start) & !is.na(arc2_touch_start))) %>%
  group_by(pcode, chirps_start) %>%
  filter(!(n() > 1 & is.na(arc2_center_start) & !is.na(chirps_start))) %>%
  ungroup()

# get simple ADM2_EN data set
adm2_df <- map(list(
  arc2_center_df,
  arc2_touch_df,
  chirps_df
), ~ select(.x, pcode, ADM2_EN)) %>%
  bind_rows() %>%
  distinct()

# add ADM2_EN to main data frame
arc2_chirps_df <- arc2_chirps_df %>%
  left_join(adm2_df, by = "pcode")
```

### Simple comparison

At a basic level, we detect about 1/3 more dry spells using ARC2 data than using CHIRPS data,
with `r nrow(arc2_center_df)` and `r nrow(arc2_touch_df)` detected with the centroid and touching methods
respectively, and only `r nrow(chirps_df)` using CHIRPS. Let's compare how and when
dry spells are detected by each method.

```{r detection-matrix}
arc2_chirps_df <- mutate(arc2_chirps_df,
  chirps_ds = !is.na(chirps_confirmation),
  arc2_centroid_ds = !is.na(arc2_center_confirmation),
  arc2_touch_ds = !is.na(arc2_touch_confirmation)
)

arc2_chirps_df %>%
  select(CHIRPS = chirps_ds,
         `ARC2 (Centroid)` = arc2_centroid_ds,
         `ARC2 (Touching)` = arc2_touch_ds) %>%
  group_by_all() %>%
  summarize(Occurrences = n(),
            .groups = "drop") %>%
  mutate(across(-Occurrences, ~ ifelse(.x, "Detected", "Not detected"))) %>%
  arrange(desc(Occurrences)) %>%
  knitr::kable()
```

We can see above a breakdown of dry spell detection by method. Interestingly, where
CHIRPS detects a dry spells, both ARC2 methods either also detect one or don't.
We should be comfortable just comparing CHIRPS dry spells to a
single ARC2 method given these similarities.

```{r onset-comparison, echo = FALSE}
arc2_chirps_df %>%
  filter(!is.na(chirps_start), !is.na(arc2_center_start)) %>%
  mutate(center_chirps = chirps_start - arc2_center_start,
         touch_chirps = chirps_start - arc2_touch_start) %>%
  arrange(touch_chirps) %>%
  mutate(row_id = row_number()) %>%
  pivot_longer(c(center_chirps, touch_chirps),
               names_to = "arc2_method",
               values_to = "day_diff") %>%
  mutate(arc2_method = ifelse(arc2_method == "touch_chirps",
                              "Centroid / Touching",
                              "Centroid only")) %>%
  ggplot(aes(x = row_id, y = day_diff, color = arc2_method)) +
  geom_point(position = "dodge", size = 3) +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  labs(y = "Days since ARC2 detection",
       title = "CHIRPS dry spell detection relative to ARC2 dry spell, days between",
       subtitle = "For dry spells detected using both CHIRPS and ARC2") +
  scale_color_manual(values = c("Centroid / Touching" = "black", "Centroid only" = "grey"))
```

From above, we can see that the dry spell detection using CHIRPS always detects the dry
spell **after** it has been detected by ARC2. This is reassuring news for the switch
to ARC2 for the observational trigger. However, what does it look like where
CHIRPS has data and ARC2 is missing it?

### Geographic comparison

```{r geo-compare}
arc2_missing_plot <- arc2_chirps_df %>%
  filter(chirps_ds, !arc2_touch_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using CHIRPS") +
  theme(plot.title.position = "plot")

chirps_missing_plot <- arc2_chirps_df %>%
  filter(!chirps_ds, arc2_touch_ds, arc2_centroid_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using ARC2",
       subtitle = "(both centroid and touching method)") +
  theme(plot.title.position = "plot")

plot_grid(arc2_missing_plot, chirps_missing_plot)
```

```{r geo-compare-arc2}
centroid_missing_plot <- arc2_chirps_df %>%
  filter(arc2_centroid_ds, !arc2_touch_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using centroids") +
  theme(plot.title.position = "plot")

touch_missing_plot <- arc2_chirps_df %>%
  filter(!arc2_centroid_ds, arc2_touch_ds) %>%
  group_by(pcode) %>%
  summarize(count = n()) %>%
  right_join(adm2, by = c("pcode" = "ADM2_PCODE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = count)) +
  scale_fill_gradient(na.value = "white",
                      breaks = 1:5) +
  geom_sf(data = adm1, fill = NA, size = 0.75) +
  coord_sf(datum = NA) +
  theme_minimal() +
  labs(title = "Dry spells only detected using touching") +
  theme(plot.title.position = "plot")

plot_grid(centroid_missing_plot, touch_missing_plot)
```

### Trigger activation

Let's consider if we're using these datasets for trigger activation and how they would
compare. We will only consider dry spells that are confirmed at some point in January
and February. Let's just quickly look at the number of dry spells detected this time period
for each method to start.

```{r trigger-chirps}
chirps_df %>%
  filter(month(chirps_confirmation) <= 2,
         str_detect(pcode, "MW3")) %>%
  mutate(trigger_year = year(chirps_confirmation)) %>%
  group_by(trigger_year) %>%
  summarize(n_triggers = n(),
            .groups = "drop")
```

```{r trigger-center}
arc2_center_df %>%
  filter(month(arc2_center_confirmation) <= 2,
         str_detect(pcode, "MW3")) %>%
  mutate(trigger_year = year(arc2_center_confirmation)) %>%
  group_by(trigger_year) %>%
  summarize(n_triggers = n(),
            .groups = "drop")
```

```{r trigger-touch}
arc2_touch_df %>%
  filter(month(arc2_touch_confirmation) <= 2,
         str_detect(pcode, "MW3")) %>%
  mutate(trigger_year = year(arc2_touch_confirmation)) %>%
  group_by(trigger_year) %>%
  summarize(n_triggers = n(),
            .groups = "drop")
```

Given that the 2 ARC2 methods are so similar, we likely only need to look at the centroid method.
This is because based on the matrix above, there are only 6 dry spells detected through
the touch method not detected through the centroid method, but all but one of those
is a city area. Alternatively, the centroid method would result in 8 triggers not captured
by the touch method. However, only one of these is an urban area and the other 7
are south-central districts in the southern region.

```{r get-rid-of-touch}
arc2_chirps_df %>%
  filter(arc2_touch_ds, !arc2_centroid_ds) %>%
  transmute(ADM2_EN, trigger_year = year(arc2_touch_confirmation))
```

```{r keep-that-centroid}
arc2_chirps_df %>%
  filter(!arc2_touch_ds, arc2_centroid_ds) %>%
  transmute(ADM2_EN, trigger_year = year(arc2_center_confirmation))
```

Although the centroid method fails to capture Zomba City (due to a lack of raster
centroid for ARC2 data), it seems sufficient to consider vs. CHIRPS excluding the
touching raster method.

### Daily value comparisons

First, let's do some 2d histograms to examine the distribution of various calculation methods.
What do we see between the centroid and touching raster methods applied?

```{r 2d-histo-arc}
daily_df %>%
  ggplot(aes(x = arc2_centroid,
             y = arc2_touch)) +
  stat_bin2d(aes(fill = log(..density..)),
             bins = 200) +
  theme_bw() +
  geom_abline(intercept = 0) +
  theme(legend.position = "none") +
  labs(title = "2D histogram, log scale",
       x = "ARC2 (centroid) precipitation",
       y = "ARC2 (touching) precipitation")
```

While it looks like the touching method might skew higher (if just slightly) at
higher levels of rainfall, we are more concerned at their behavior at lower levels.

```{r 2d-histo-arc-1}
daily_df %>%
  filter(arc2_centroid <= 1 | arc2_touch <= 1) %>%
  ggplot(aes(x = arc2_centroid,
             y = arc2_touch)) +
  geom_bin2d(bins = 50,
             aes(fill = log(..density..))) +
  theme_bw() +
  geom_abline(intercept = 0) +
  theme(legend.position = "none") +
  labs(title = "2D histogram, log scale",
       subtitle = "Only dates where at least one metric measures <= 0.2mm",
       x = "ARC2 (centroid) precipitation",
       y = "ARC2 (touching) precipitation")
```

And above we can see that the centroid method in general trends much higher at these levels.
This intuitively makes sense as the touching method takes the average of more raster
cells than the centroid method, so will tend to regress toward the mean at the extremes.
Anyway, given their similar results in identifying dry spells and the results above,
should be fine to simply compare the centroid method against CHIRPS and not the
touching method.

```{r 2d-histo-chirps}
daily_df %>%
  ggplot(aes(x = arc2_centroid,
             y = chirps)) +
  geom_bin2d(bins = 200,
             aes(fill = log(..density..))) +
  theme_bw() +
  geom_abline(intercept = 0) +
  theme(legend.position = "none") +
  labs(title = "2D histogram, log scale",
       x = "ARC2 (centroid) precipitation",
       y = "CHIRPS precipitation")
```

```{r 2d-histo-chirps-2}
daily_df %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2)%>%
  ggplot(aes(x = arc2_centroid,
             y = chirps)) +
  geom_bin2d(bins = 200,
             aes(fill = log(..density..))) +
  theme_bw() +
  geom_abline(intercept = 0) +
  theme(legend.position = "none") +
  labs(title = "2D histogram, log scale",
       subtitle = "Only dates where at least one metric measures <= 0.2mm",
       x = "ARC2 (centroid) precipitation",
       y = "CHIRPS precipitation")
```

The above graph is obviously difficult to interpret given some of the extreme values.
We could limit the axes, but might be more intuititive to graph instead the differences.

```{r diff-chirps}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       x = "Chirps - ARC2",
       y = "Count")
```

```{r diff-chirps-2}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       subtitle = "Only dates where at least one metric measures <= 0.2mm",
       x = "Chirps - ARC2",
       y = "Count")
```

```{r diff-chirps-excl0}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 6000)) +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       subtitle = "Only dates where at least one metric measures <= 0.2mm; zoomed",
       x = "Chirps - ARC2",
       y = "Count")
```
### January / February comparisons

Since our action trigger will be monitored only in January and February 2022,
important to validate the comparability of ARC2 and CHIRPS across that period
solely to ensure the broader trends still hold true (which indicate that ARC2 is
as sufficient if not preferable for dry spell detection).

```{r diff-chirps-feb}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  filter(lubridate::month(date) <= 2) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       subtitle = "January and February only",
       x = "Chirps - ARC2",
       y = "Count")
```

```{r diff-chirps-2-feb}
daily_df %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  filter(arc2_centroid <= 0.2 | chirps <= 0.2,
         lubridate::month(date) <= 2) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       subtitle = "January and February only; only dates where at least one metric measures <= 0.2mm",
       x = "Chirps - ARC2",
       y = "Count")
```
We can again see that CHIRPS data in this period does indeed follow the same trend.

### During mismatched dry spells

Another final piece of checking to do would be to see what is going on with
daily values in the periods where one method detects a dry spell but the other
detects no overlapping dry spell at all.

```{r arc2-no-spell}
arc2_no_spell_daily <- arc2_chirps_df %>%
  filter(chirps_ds, !arc2_centroid_ds) %>%
  fuzzy_inner_join(
    daily_df,
    match_fun = list(`==`, `<=`, `>=`),
    by = c(
      "pcode" = "pcode",
      "chirps_start" = "date",
      "chirps_end" = "date"
    )) %>%
  select(pcode = pcode.x,
         date,
         chirps,
         arc2_centroid)

arc2_no_spell_daily %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       subtitle = "Only dates where CHIRPS detected a dry spell but none was detected by ARC2",
       x = "Chirps - ARC2",
       y = "Count")
```

Clearly, by defining our subset, we see many instances where the CHIRPS measured
precipitation is lower than ARC2. But where is actually coming about?

```{r arc2-missing-why}
arc2_no_spell_daily %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid,
         chirps_type = ifelse(chirps > 0, "> 0mm", "0mm"),
         ) %>%
  group_by(chirps_type) %>%
  summarize(`ARC2 0mm` = sum(arc2_centroid == 0, na.rm = T),
            `ARC2 > 0mm` = sum(arc2_centroid > 0, na.rm = T),
            `ARC2 > 0mm & > chirps` = sum(arc2_centroid > 0 & arc2_centroid > chirps, na.rm = T))
```
From the above, it's clear that the vast majority of the time, ARC2 and CHIRPS align,
just that there are instances where CHIRPS measures 0 precipitation but ARC2 measures
some, and another 8 where both ARC2 and CHIRPS measure precipitation but CHIRPS
measures none.

```{r arc2-ts-graph}
arc2_no_spell_daily %>%
  filter(!is.na(arc2_centroid)) %>%
  mutate(g = paste(pcode, lubridate::year(date), sep = ", ")) %>%
  group_by(g) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(id = row_number()) %>%
  ggplot(aes(x = id)) +
  geom_line(aes(y = chirps), color = "grey") +
  geom_line(aes(y = arc2_centroid)) +
  facet_wrap(~g, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(x = "Days of dry spell",
       y = "Precipitation",
       title = "Precipitation values of ARC2 and CHIRPS during CHIRPS dry spells",
       subtitle = "ARC2 values in black")
```
Aha! Very informative! It's quite clear that in all but one of the dry spells,
ARC2 is measuring "extreme" spikes in precipitation surrounding primarily days with
none. In some instances, like MW209 or MW203, these are spikes also measured by CHIRPS, just
with higher precipitation recorded by ARC2. In the majority, however, the spikes 
are registered solely by ARC2, whereas CHIRPS records little to no precipitation.

### Flip it!

Just for a bit of extra exploration, let's quickly run the same code to assess
CHIRPS daily values where CHIRPS doesn't detect an overlapping dry spell when
ARC2 detects one.

```{r chirps-no-spell}
chirps_no_spell_daily <- arc2_chirps_df %>%
  filter(!chirps_ds, arc2_centroid_ds) %>%
  fuzzy_inner_join(
    daily_df,
    match_fun = list(`==`, `<=`, `>=`),
    by = c(
      "pcode" = "pcode",
      "arc2_center_start" = "date",
      "arc2_center_end" = "date"
    )) %>%
  select(pcode = pcode.x,
         date,
         chirps,
         arc2_centroid)

chirps_no_spell_daily %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid) %>%
  ggplot(aes(x = chirps_arc2_diff)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(title = "Precipitation difference, CHIRPS - ARC2",
       subtitle = "Only dates where CHIRPS detected a dry spell but none was detected by ARC2",
       x = "Chirps - ARC2",
       y = "Count")
```

```{r chirps-missing-why}
chirps_no_spell_daily %>%
  mutate(chirps_arc2_diff = chirps - arc2_centroid,
         arc2_type = ifelse(arc2_centroid > 0, "> 0mm", "0mm"),
         ) %>%
  group_by(arc2_type) %>%
  summarize(`CHIRPS 0mm` = sum(chirps == 0, na.rm = T),
            `CHIRPS > 0mm` = sum(chirps > 0, na.rm = T),
            `CHIRPS > 0mm & > chirps` = sum(chirps > 0 & chirps > arc2_centroid, na.rm = T))
```

```{r chirps-ts-graph}
chirps_no_spell_daily %>%
  mutate(g = paste0(pcode, lubridate::year(date))) %>%
  group_by(g) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(id = row_number()) %>%
  ggplot(aes(x = id)) +
  geom_line(aes(y = chirps)) +
  geom_line(aes(y = arc2_centroid), color = "grey") +
  facet_wrap(~g, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(x = "Days of dry spell",
       y = "Precipitation",
       title = "Precipitation values of ARC2 and CHIRPS during ARC2 dry spells",
       subtitle = "CHIRPS values in black")
```

Contrary to ARC2 values were detecting dry spells