---
title: '<img src="https://drive.google.com/uc?id=118y5T73-iSUZoAMtoJGddxq9QzD_GDKX" style="height:40px;float:left; margin:10px" /><img src="https://drive.google.com/uc?id=1fHQUzF3ZjaoHj9KQ33-94dK_X1hcmjzW" style="height:50px;float:left; margin:10px" />'
pagetitle: 'ETH Baseline Risk Analysis' # Change title to something appropriate for your work
output:
  html_document:
    css: ../../country_template/docs/style.css
    includes:
      in_header: ../../country_template/docs/header.html
    df_print: paged
    toc: yes
    toc_float: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: inline
---

<br> 
<br>
<br>

# Ethiopia Baseline Risk Analysis
Analysis last updated 06-10-2021.

<br>

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
# source("") # Link to external script
library(kableExtra)
library(knitr)
source('make_plots.R')
knitr::opts_chunk$set(echo = FALSE) # do not print code by default
knitr::opts_chunk$set(include = FALSE) # do not print output by default
```


This document is an initial exploration into risk of humanitarian need in Ethiopia. We're focusing most closely on climate-related risks, such as floods and drought. This document presents the outputs of basic data collection and analysis. It is intended to provide a starting point for a more in depth risk analysis for Ethiopia's upcoming Humanitarian Needs Overview. 

## Historical Analysis

Relevant datasets include: 

1. [Global Flood Database](http://global-flood-database.cloudtostreet.info/#interactive-map)
- Gridded data of historical flood extents available via [Google Earth Engine](https://developers.google.com/earth-engine/datasets/catalog/GLOBAL_FLOOD_DB_MODIS_EVENTS_V1).
- Includes flood events from 2000-2018. Ethiopia has 28 events recorded.
- Events identified from the Dartmouth Flood Observatory, with flood extent mapped using MODIS satellite imagery.

2. [EM-DAT](https://www.emdat.be/database)
- Historical catalogue of both flood and drought events from 1965-2021.
- Dataset includes 74 events.
- Geolocated somewhat inconsistently to most often the Admin 1 or 2 level. 

The figure below shows the national count of shock events per year, as recorded in EM-DAT. 

```{r ts-basic, dpi=300, include=T, message= FALSE, warning=FALSE, fig.height=2, fig.align="center", fig.cap='National count of shock events per year'}

# Put a figure in here, for example 

plot_disaster_annual

```

The figure below shows the frquency of floods (left) and drought (right) by region since 1965. NOTE that the geolocation of events in these figures still needs to be fully validated.

```{r historical-freq, dpi=300, include=T, message= FALSE, warning=FALSE, fig.height=5, fig.show="hold", fig.cap = 'Historical frequency of shock events by Region', out.width="50%"}

# Put a figure in here, for example 

flood_map

drought_map

```

The tables below show the full lists of both the flood and drought events from EM-DAT. 

```{r event-table-flood, dpi=300, include=T, message= FALSE, warning=FALSE}

emdat_flood %>%
  select(c(Year, Location, Origin, `Total Affected`)) %>%
  knitr::kable(caption = "List of flood events",
               align = c('l')) %>%
        collapse_rows(columns = 1, valign = "top") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                      fixed_thead = T) %>%
        scroll_box(width = "100%", height = "500px")

```

<br>
<br>

```{r event-table-drought, dpi=300, include=T, message= FALSE, warning=FALSE}

emdat_drought %>%
  select(c(Year, Location, Origin, `Total Affected`)) %>%
  knitr::kable(caption = "List of drought events",
               align = c('l')) %>%
        collapse_rows(columns = 1, valign = "top") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                      fixed_thead = T) %>%
        scroll_box(width = "100%", height = "500px")

```

<br>
<br>

3. [IDMC](https://data.humdata.org/dataset/idmc-idp-data-for-ethiopia)
- Number of IDPs, relating to conflict and disaster.
- Not subnational.
- Data available between 2008-2020.

4. [IOM DTM Site Assessment](https://data.humdata.org/dataset/ethiopia-baseline-assessment-data-iom-dtm)
- Assessment July 2021.
- At Admin 3 level.
- For each site, includes reason for displacement and breakdown of population by age and gender.

The figures below show the basic spatial distribution, volume, and demographics of Ethiopia's IDP population as identified by the IOM DTM Site Assessment conducted in July 2021. 

```{r img1, echo = FALSE, include=T, fig.cap='IDP population breakdown by gender at Admin 2 level, symbols scaled by total IDP population', out.extra='style=padding:5px;"', fig.align='center'}
images1 <- c('gender-map-pop.png')
include_graphics(images1)
```

```{r img2, echo = FALSE, include=T, fig.cap='IDP population breakdown by age at Admin 2 level, symbols scaled by total IDP population', out.extra='style=padding:5px;"', fig.align='center'}
images2 <- c('age-map-pop.png')
include_graphics(images2)
```

```{r img3, echo = FALSE, include=T, fig.cap='Reason for displacement of IDPs at Admin 2 level', out.extra='style=padding:5px;"', fig.align='center'}
images3 <- c('reason-map-pop.png')
include_graphics(images3)
```

5. [FAOSTAT Food Security Indicators](https://data.humdata.org/dataset/faostat-food-security-indicators-for-ethiopia)
- National level indicators from 2000-2020
- Such as: Prevalence of undernourishment (percent) (3-year average), Number of severely food insecure people (million) (3-year average), Percentage of population using safely managed sanitation services (Percent)

## Near-present situation

1. [Standardized Precipitation Index](https://code.earthengine.google.com/5d3600853925e070e57c9f733072e274)
- Researched in past MapAction work in Ethiopia: 

> Standardized Precipitation Index (SPI) is the most commonly used indicator worldwide for detecting and characterizing meteorological droughts. It measures precipitation anomalies at a given location based on a comparison of observed total precipitation amounts for an accumulation period of interest (e.g. 3, 6, 9 months), with the long-term historic rainfall record for that period. Rainfall deficits (i.e. meteorological droughts) are indicated as SPI decreases below -1.0, while excess rainfall are indicated as SPI increases above 1.0. As SPI values are in units of standard deviation from the long-term mean, the indicator can be used to compare precipitation anomalies for any geographic location and for any number of time scales. The World Meteorological Organization has recommended that the SPI be used by all National Meteorological and Hydrological Services around the world to characterize meteorological droughts


